Maximum Entropy Prediction Markets

### 2014-02-10 WriteUp/maxentpredmkts.tex
adding more notes and some fixes

\subsection{Related Work}
\jake{I'll add more here. Sebastien: do you have anything to put in here regarding the exponential utility stuff?}
Designing prediction markets to handle a large outcome space is an active area of research.  In \cite{Chen08}, the authors use a restricted betting language to design efficient markets for a combinatorial outcome space. This technique is generalized by \cite{Pennock11}. \cite{Gao09} consider extending various automated market makers to an infinite outcome space. For the logarithmic market scoring rule they show that unbounded market maker loss can result in this setting.   \cite{Abernethy11} and \cite{OthmanS11} specify frameworks under which they design cost function based markets that satisfy the desirable property of bounded market maker loss even in infinite outcome spaces.

@@

If wealth is distributed according to a probability measure $P$, then the \emph{certainty equivalent} of a random amount of wealth is defined as
$$
\CE(w) = U_a^{-1} \Exp_P\left[ U_a(w) \right].
$$
Suppose as before that the agent's belief over outcomes takes the form of a density $p$ with respect to base measure $\nu$. There is a close relationship between the log-partition function and the certainty equivalent under exponential utility.
%
\begin{lemma} \label{lem:exp-util}
The certainty equivalent of the agent's expected profit, under exponential utility, when acquiring shares $\delta$ under a market state of $\theta$ is
\begin{equation}
\log a - T_p(-a\delta) - aC(\theta+\delta) + aC(\theta),
\end{equation}
where $T_p$ is the log partition function~(\ref{log-part}) with a base measure of $p\,d\nu$. Furthermore, if the agent's belief is an exponential family with parameter $\htheta$, we have
%
\begin{equation*}
T_p(\delta) = T(\htheta + \delta) - T(\htheta),
\end{equation*}
%
where $T$ is the usual log partition function with base measure $\nu$.
\end{lemma}
%
%
\begin{proof}
Explicitly, the certainty equivalent of the profit is
%
\begin{eqnarray*}
& & \CE(\,\la\delta,\phi(x)\ra - [C(\theta+\delta) - C(\theta)]\,)\\
& = & -\log \int_{\mX} \frac{1}{a} \exp\left( \la-a\delta,\phi(x)\ra + a [C(\theta+\delta) - C(\theta)]  \right) p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - \log \int_{\mX} \exp\left( \la-a\delta,\phi(x)\ra \right)  p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - T_p(-a\delta).
\end{eqnarray*}
%
For the second part of the result, we have
%
\begin{eqnarray*}
T_p(\delta) & = & \log \int_{\mX} \exp(\la\delta,\phi(x)\ra)\,p(x;\htheta)\, d\nu(x)\\
& = &  \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta+\delta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} p(x;\htheta+\delta) \, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta),
\end{eqnarray*}
%
where the last line follows from the fact that density $p(x;\htheta+\delta)$ integrates to 1.
\end{proof}

@@

\subsection{Reinterpreting repeated trades}
In previous sections we have analyzed trader behavior as if it is his first entry into the market. In this section we will quantify how prior exposure in the market affects a trader's choices. In particular, we will show that a trader who has previously purchased shares in a market will, on subsequent entry, behave as if this purchase has updated his private belief. This result implies that any financial trade made in a market is equivalent to changing the trader's effective belief.

As before, we consider the exponential family prediction market where traders have exponential belief. Let us also suppose that traders in this market have exponential utility \[
U(w) = -\frac{1}{a} \exp(-aw).
\]
%
Here $a$ is the coefficient of risk aversion (higher means more risk averse, and the utility function is more concave).

@@

\subsection{Equilibrium Market State for Exponential Utility Agents}
We have shown that every exponential-utility maximizing trader picks the share vector $\delta$ so that the eventual market state can be represented as a convex combination of the initial market state and the natural parameter of his (exponential family) belief distribution. In this section we will compute the equilibrium state in an exponential family market with multiple exponential-utility maximizing traders.

Recall the following result from game theory. %need a reference/name for this result?
\begin{thm}\label{thm:potl}
Let $U_{i}(\vec{\delta})$ be the utility function of the $i^{th}$ trader given strategies $\vec{\delta}\defeq \delta_{1},\ldots,\delta_{i},\ldots,\delta_{n}$. If there exists a potential function $g(\vec{\delta})$ such that $$U_{i}(\vec{\delta})-U_{i}(\vec{\delta}_{-i},\delta_{i}')=g(\vec{\delta})-g(\vec{\delta}_{-i},\delta_{i}')$$
then when $g(\vec{\delta})$ is maximized, $\vec{\delta}$ is an equilibrium.
\end{thm}¥
In the exponential family market, the cost function $C$ is identical to the log partition function $T$.  Let $\vec{\delta}$ be the vector of vectors of shares purchased by every trader in the market when the market has reached equilibrium, $\theta$ the initial market state, $\htheta_{i}$ the natural parameter of trader $i$'s belief distribution and $a_{i}$ his utility parameter.

Define a potential function as  $$g(\vec{\delta})\defeq  T(\theta+\sum_{i}\delta_{i})+\sum_{i}\frac{1}{a_{i}}T(\htheta_{i}-a_{i}\delta_{i})$$

%is the following expression for utility correct?
Now the utility of trader $i$ is $U_{i}(\vec{\delta})=- T(\theta+\sum_{j}\delta_{j}) + T(\theta+\sum_{j\neq i}\delta_{j}) - \frac{1}{a_{i}} T(\htheta_{i} - a_{i}\delta_{i}) + \frac{1}{a_{i}} T(\htheta_{i})$.
Thus, Theorem \ref{thm:potl} applies and we can find the equilibrium market state by maximizing $g(\vec{\delta})$ for each $\delta_{i}$.
\begin{eqnarray*}
\nabla_{\delta_{i}}g(\vec{\delta})&=&\nabla T(\theta+\sum_{j=1}^{n}\delta_{i})-\nabla T(\htheta_{i}-a_{i}\delta_{i})\\
&=&0
\end{eqnarray*}¥
This can be achieved by equating the arguments. That is, for each trader $i$, 
\begin{equation}\label{eq:eqbm}
\htheta_{i}-a_{i}\delta_{i}=\theta+\sum_{j=1}^{n}\delta_{j}
\end{equation}¥
Rewriting, we have for each trader $i$, 
$$\frac{\htheta_{i}}{a_{i}}-\delta_{i}=\frac{1}{a_{i}}(\theta+\sum_{j=1}^{n}\delta_{j})$$
Thus,
$$\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)-\sum_{i=1}^{n}\delta_{i}=\left(\theta+\sum_{j=1}^{n}\delta_{j}\right)\sum_{i=1}^{n}\frac{1}{a_{i}}$$
And
$$\sum_{j=1}^{n}\delta_{j}=
\frac{\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)-\theta\sum_{i=1}^{n}\left(\frac{1}{a_{i}}\right)}{1+\sum_{i=1}^{n}\frac{1}{a_{i}}}$$
Substituting in Equation \ref{eq:eqbm} we have the following expression for the final market state.
$$\theta+\sum_{j=1}^{n}\delta_{j}=
\frac{\theta+\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)}{1+\sum_{i=1}^{n}\frac{1}{a_{i}}}$$
Thus the equilibrium state is a convex combination of trader beliefs and initial market state.

>>>>

\jake{I want to add more to this list. It's ok but doesn't totally reflect what's in the last few sections.}
\subsection{Related Work}
\jake{I'll add more here. Sebastien: do you have anything to put in here regarding the exponential utility stuff?}
Designing prediction markets to handle a large outcome space is an active area of research.  In \cite{Chen08}, the authors use a restricted betting language to design efficient markets for a combinatorial outcome space. This technique is generalized by \cite{Pennock11}. \cite{Gao09} consider extending various automated market makers to an infinite outcome space. For the logarithmic market scoring rule they show that unbounded market maker loss can result in this setting.   \cite{Abernethy11} and \cite{OthmanS11} specify frameworks under which they design cost function based markets that satisfy the desirable property of bounded market maker loss even in infinite outcome spaces.

@@

\jake{This set of ideas is nice and needs to be sold more strongly in the intro} If wealth is distributed according to a probability measure $P$, then the \emph{certainty equivalent} of a random amount of wealth is defined as
$$
\CE(w) = U_a^{-1} (\Exp_P\left[ U_a(w) \right]).
$$
Suppose as before that the agent's belief over outcomes takes the form of a density $p$ with respect to base measure $\nu$. There is a close relationship between the log-partition function and the certainty equivalent under exponential utility.
%
\begin{lemma} \label{lem:exp-util}
The certainty equivalent of the agent's expected profit, under exponential utility, when acquiring shares $\delta$ under a market state of $\theta$ is
\begin{equation}
\log a - T_p(-a\delta) - aC(\theta+\delta) + aC(\theta),
\end{equation}
where $T_p$ is the log partition function~(\ref{log-part}) with a base measure of $p\,d\nu$. Furthermore, if the agent's belief is an exponential family with parameter $\htheta$, we have
%
\begin{equation*}
T_p(\delta) = T(\htheta + \delta) - T(\htheta),
\end{equation*}
%
where $T$ is the usual log partition function with base measure $\nu$.
\end{lemma}
%
%
\begin{proof}
Explicitly, the certainty equivalent of the profit is
%
\begin{eqnarray*}
& & \CE(\,\la\delta,\phi(x)\ra - [C(\theta+\delta) - C(\theta)]\,)\\
& = & -\log \int_{\mX} \frac{1}{a} \exp\left( \la-a\delta,\phi(x)\ra + a [C(\theta+\delta) - C(\theta)]  \right) p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - \log \int_{\mX} \exp\left( \la-a\delta,\phi(x)\ra \right)  p(x)d\nu(x)\\
& = & \log a - a [C(\theta+\delta) - C(\theta)] - T_p(-a\delta).
\end{eqnarray*}
%
For the second part of the result, we have\jake{Is the $\hat \cdot$ missing from the computations below?}
%
\begin{eqnarray*}
T_p(\delta) & = & \log \int_{\mX} \exp(\la\delta,\phi(x)\ra)\,p(x;\htheta)\, d\nu(x)\\
& = &  \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} \exp( \la\delta+\htheta,\phi(x)\ra - T(\htheta+\delta) )\, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta) + \log \int_{\mX} p(x;\htheta+\delta) \, d\nu(x)\\
& = &  T(\htheta+\delta) - T(\theta),
\end{eqnarray*}
%
where the last line follows from the fact that density $p(x;\htheta+\delta)$ integrates to 1.
\end{proof}

@@

\subsection{Repeated Trading and the Effective Belief}
\jake{I need to clean up a few things in this section.}
In previous sections we have analyzed trader behavior as if it is his first entry into the market. In this section we will quantify how prior exposure in the market affects a trader's choices. In particular, we will show that a trader who has previously purchased shares in a market will, on subsequent entry, behave as if this purchase has updated his private belief. This result implies that any financial trade made in a market is equivalent to changing the trader's effective belief.

% As before, we consider the exponential family prediction market where traders have exponential belief. Let us also suppose that traders in this market have exponential utility \[
% U(w) = -\frac{1}{a} \exp(-aw).
% \]
% %
% Here $a$ is the coefficient of risk aversion (higher means more risk averse, and the utility function is more concave).

@@

\subsection{Equilibrium Market State for Exponential Utility Agents}
\jake{need to mention the connection to potential games in the intro, it's a nice little extra bit.}
We have shown that every exponential-utility maximizing trader picks the share vector $\delta$ so that the eventual market state can be represented as a convex combination of the initial market state and the natural parameter of his (exponential family) belief distribution. In this section we will compute the equilibrium state in an exponential family market with multiple exponential-utility maximizing traders.

We now review a well-known result from game theory regarding the class of \emph{potential games} \cite{monderer1996potential}. Note that we say a function $f(\vec{x})$ is at a \emph{local optimum} if changing any coordinate of $\vec{x}$ does not increase the value of $f$.
\begin{thm}\label{thm:potl}
Let $U_{i}(\vec{\delta})$ be the utility function of the $i^{th}$ trader given strategies $\vec{\delta}\defeq (\delta_{1},\ldots,\delta_{i},\ldots,\delta_{n})$. If there exists a potential function $\Phi(\vec{\delta})$ such that $$U_{i}(\vec{\delta})-U_{i}(\vec{\delta}_{-i},\delta_{i}')=\Phi(\vec{\delta})-\Phi(\vec{\delta}_{-i},\delta_{i}')$$
then $\vec{\delta}$ is a Nash equilibrium if and only if $\Phi(\vec{\delta})$ is at a local optimum.
\end{thm}¥
In the exponential family market, the cost function $C$ is identical to the log partition function $T$.  Let $\vec{\delta}$ be the vector of vectors of shares purchased by every trader in the market when the market has reached equilibrium, $\theta$ the initial market state, $\htheta_{i}$ the natural parameter of trader $i$'s belief distribution and $a_{i}$ his utility parameter.

Define a potential function as  $$\Phi(\vec{\delta})\defeq  T(\theta+\sum_{i}\delta_{i})+\sum_{i}\frac{1}{a_{i}}T(\htheta_{i}-a_{i}\delta_{i})$$

%is the following expression for utility correct?
Now the utility of trader $i$ is $U_{i}(\vec{\delta})=- T(\theta+\sum_{j}\delta_{j}) + T(\theta+\sum_{j\neq i}\delta_{j}) - \frac{1}{a_{i}} T(\htheta_{i} - a_{i}\delta_{i}) + \frac{1}{a_{i}} T(\htheta_{i})$.
Thus, Theorem \ref{thm:potl} applies and we can find the equilibrium market state by maximizing $\Phi(\vec{\delta})$ for each $\delta_{i}$.
\begin{eqnarray*}
\nabla_{\delta_{i}}\Phi(\vec{\delta})&=&\nabla T(\theta+\sum_{j=1}^{n}\delta_{i})-\nabla T(\htheta_{i}-a_{i}\delta_{i}) = 0
\end{eqnarray*}¥
This can be achieved by equating the arguments. That is, for each trader $i$, 
\begin{equation}\label{eq:eqbm}
\htheta_{i}-a_{i}\delta_{i}=\theta+\sum_{j=1}^{n}\delta_{j}
\end{equation}¥
Rewriting, we have for each trader $i$, 
$$\frac{\htheta_{i}}{a_{i}}-\delta_{i}=\frac{1}{a_{i}}(\theta+\sum_{j=1}^{n}\delta_{j})$$
Thus,
$$\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)-\sum_{i=1}^{n}\delta_{i}=\left(\theta+\sum_{j=1}^{n}\delta_{j}\right)\sum_{i=1}^{n}\frac{1}{a_{i}}$$
And
$$\sum_{j=1}^{n}\delta_{j}=
\frac{\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)-\theta\sum_{i=1}^{n}\left(\frac{1}{a_{i}}\right)}{1+\sum_{i=1}^{n}\frac{1}{a_{i}}}$$
Substituting in Equation \ref{eq:eqbm} we have the following expression for the final market state.
$$\theta+\sum_{j=1}^{n}\delta_{j}=
\frac{\theta+\sum_{i=1}^{n}\left(\frac{\htheta_{i}}{a_{i}}\right)}{1+\sum_{i=1}^{n}\frac{1}{a_{i}}}$$
Thus the equilibrium state is a convex combination of trader beliefs and initial market state.
